Synthesis of Linear Functional Programs

### 2021-05-08 article/article.tex
slow work

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  Regardless of the kind
of specification, program synthesizers have to deal with two main
inherent sources of complexity: searching over a vast space of
(potentially) valid programs, and interpreting user intent.
%
In this work we will explore type-based synthesis where specifications
take the form of linear types, which are types that limit resource
usage in programs.  Concretely, we will explore the proof-theoretic
foundations of linear types via the Curry-Howard correspondence with
linear logic, viewing program synthesis under the lens of (linear
logic) bottom-up proof search.
%

@@

\mypara{Type-Based Synthesis}
% Type based synthesis
Type-based synthesis uses types as a form of program specification,
automatically or semi-automatically producing a program expression
with the given type and so matching the specification.
%
In order to make type-based specifications more expressive, type-based
synthesis frameworks use rich type systems such as refinement
types~\cite{} and polymorphic types~\cite{}.
\todo[inline]{[Estive a procurar e fiquei
um pouco confuso. polimorfismo é rich types?  dependent types are rich
types? aqueles constraints são os dependent types pelo que
percebi... mas n me parece correto. arithmetic types? temos de
rever. Por agora vou por apenas os tópicos que o professor escreveu]
-- Não há propriamente uma definição rigorosa de ``rich types'',
geralmente tudo o que vá além do tipico pode ser qualificado como
rico.}

@@

%
This work explores the process of synthesizing linear functional
programs from types based on linear logic propositions (i.e.~linear
types) by leveraging the Curry-Howard correspondence.  The
correspondence states that propositions in a logic have a direct
mapping to types, and well-typed programs correspond to to proofs of
those propositions.  As such, having a type be a proposition in linear
logic, we can relate a proof of that proposition directly to a linear
functional program — finding a proof is finding a program with that
type. Thus, we formulate synthesis as proof search in linear logic,
which allows us to inform our work with approaches from the proof
search literature.  Linear types differ from more traditional types in
that they constrain resource usage in programs by \emph{statically}
limiting the number of times certain resources can be used during
their lifetime.  They can be applied to resource-aware programming
such as concurrent programming (e.g. session types for message passing
concurrency~\cite{}), and to memory-management (e.g.~Rust's ownership
types).

@@

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{} that captures \dots. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting the
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, the \dots \todo{Identity function!}

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus, and their proofs are
exactly the \dots.
The main observations made are ``propositions as types'', ``proofs as
programs'', ``proof checking as type checking''~\cite{}.

Inference in ND are categorized as introduction or elimination, these
correspond to constructors and destructors in programming languages
(e.g. function abstraction vs application, construction of a pair vs
projection).

We introduce a few typing rules to display how a type system can be
formalized.  In natural deduction, there are both \emph{introduction}
and \emph{elimination} rules for almost all logical
connectives.\todo{se mencionares intro e elim antes nao precisas de
  repetir aqui.}  The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
  \]
  When seen as a typing rule for the $\lambda$-calculus, it
  corresponds to the rule for typing variables -- variable $u$ has
  type $A$ if the typing environment contains a variable $u$ of
  type $A$.

  As another concrete example, let us consider the rule for
  implication ($A \supset B$):
\[
    \infer*[right=($\supset I$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \supset B }
  \]
  Logically, the rule states that to prove $A\supset B$, we assume $A$
  and prove $B$. Through the Curry-Howard correspondence, implication
  corresponds to the function type -- the program $\lambda u. M$ has
  type $A \supset B$, provided $M$ has type
  type $B$ under the assumption that $u$ is a variable of type $A$.  The rule
  shown above is an \emph{introduction} rule, that introduces the
  ``implication'' connective. 
%$\rightarrow$ $\Rightarrow$, $\Longrightarrow$, $\longrightarrow$

  \todo[inline]{Mesma historia para a aplicacao/modus ponens.
Um bocadinho das ``consequencias'' do Curry-Howard. }

The Curry-Howard correspondence generalizes beyond the simply-typed
$\lambda$-calculus and propositional intuitionistic logic.
\todo{Mencionar polimorfismo corresponde a logica de ordem superior,
  etc. E tambem para logica linear \dots}

@@

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.
To present the formalization of this logic, besides the new connectives, we need to introduce the \emph{resource-aware context} $\Delta$.
 In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types, but in which each and every variable must be used exactly once.
 So, to introduce the connective $\tensor$ which defines a multiplicative pair of propositions, we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
 to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
it useable more than once. To distinguish the ``infinite'' variables, a separate context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.

@@

Concrete works
\begin{itemize}

\item Type-and-example-directed
  synthesis~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
  or how to turn your type system upside down.

@@

\end{itemize}

@@

Mention that you've already implemented a type-checker for this
(useful as a prelim. exercise but also later, for \emph{validation}).

@@

\[
  \infer[$(\lolli\! I)$]
  {\Delta , x{:}A \vdash M : B }
  {\Delta \vdash \lambda x {:} A . M : A \lolli B}
  \quad
  \infer[$(\lolli\! E)$]
  {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
  {\Delta_1, \Delta_2 \vdash M\,N : B}
\]

>>>>

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  Regardless of the kind
of specification, program synthesizers have to deal with two main
inherent sources of complexity: searching over a vast space of
(potentially) valid programs, and interpreting user intent.
%
In this work we will explore type-based synthesis where specifications
take the form of linear types, which are types that limit resource
usage in programs.  Concretely, we will explore the proof-theoretic
foundations of linear types via the Curry-Howard correspondence with
linear logic, viewing program synthesis under the lens of (linear
logic) bottom-up proof search.

@@

\mypara{Type-Based Synthesis}
% Type based synthesis
Type-based synthesis uses types as a form of program specification,
automatically or semi-automatically producing a program expression
with the given type and so matching the specification.
%
In order to make type-based specifications more expressive, type-based
synthesis frameworks use rich type systems such as refinement
types~\cite{} and polymorphic types~\cite{}.

@@

%
This work explores the process of synthesizing linear functional
programs from types based on linear logic propositions (i.e.~linear
types) by leveraging the Curry-Howard correspondence.  The
correspondence states that propositions in a logic have a direct
mapping to types, and well-typed programs correspond to proofs of
those propositions.  As such, having a type be a proposition in linear
logic, we can relate a proof of that proposition directly to a linear
functional program — finding a proof is finding a program with that
type. Thus, we formulate synthesis as proof search in linear logic,
which allows us to inform our work with approaches from the proof
search literature.  Linear types differ from more traditional types in
that they constrain resource usage in programs by \emph{statically}
limiting the number of times certain resources can be used during
their lifetime.  They can be applied to resource-aware programming
such as concurrent programming (e.g. session types for message passing
concurrency~\cite{}), and to memory-management (e.g.~Rust's ownership
types).

@@

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, $\lambda x{:}A.x$, the identity function, is a functional abstraction
taking an argument of type $A$ and returning it back.

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation), their proofs are
exactly the program with that type (``proofs as programs'' observation), and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.

Inference rules in natural deduction are categorized as introduction or elimination, these
correspond to constructors and destructors in programming languages
(e.g. function abstraction vs application, construction of a pair vs
projection).

We introduce a few typing rules to display how a type system can be
formalized. The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
\]
When seen as a typing rule for the $\lambda$-calculus, it
corresponds to the rule for typing variables -- variable $u$ has
type $A$ if the typing environment contains a variable $u$ of
type $A$.

As another concrete example, let us consider the rule for
implication ($A\rightarrow~B$):
\[
    \infer*[right=($\rightarrow I$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \rightarrow B}
\]
Logically, the rule states that to prove $A\rightarrow B$, we assume $A$
and prove $B$. Through the Curry-Howard correspondence, implication
corresponds to the function type -- the program $\lambda u. M$ has
type $A \rightarrow B$, provided $M$ has type
$B$ under the assumption that $u$ is a variable of type $A$. The rule
shown above is an \emph{introduction} rule, that introduces the
``implication'' connective.

Finally, let us consider an \emph{elimination} rule, also for the implication connective:
\[
    \infer*[right=($\rightarrow E$)]
    {\Gamma \vdash M : A \rightarrow B \and \Gamma \vdash N : A}
    {\Gamma \vdash M\,N : B}
\]
Elimination rules are easier to think of in a top-down manner. Logically, this rule states
that if we prove $A \rightarrow B$ and $A$ we can prove $B$. Through the Curry-Howard correspondence,
implication elimination corresponds to function application, and its type is the function return type
-- the program $M N$ has type $B$, provided $M$ has type $A \rightarrow B$ and $N$ has type $A$.

The Curry-Howard correspondence generalizes beyond the simply-typed
$\lambda$-calculus and propositional intuitionistic logic. It extends to
the realms of polymorphism (higher-order logic \todo{review - {https://en.wikipedia.org/wiki/Lambda_cube}} second-order propositional calculus),
dependent types (first-order predicate logic), and every other extension
of natural deduction and simply-typed $\lambda$-calculus -- which also means
linear logic and linear types.

@@

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.
To present the formalization of this logic, besides the new connectives, we 
need to introduce the \emph{resource-aware context} $\Delta$.
 In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types,
 but in which each and every variable must be used exactly once.
To introduce the connective $\tensor$ which defines a multiplicative pair of propositions,
 we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
 to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
it useable more than once. To distinguish the ``infinite'' variables, a separate context is used --
$\Gamma$. $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
The linear typing judgment for the introduction of the exponential $\bang$ takes the form:
\todo{A formula está um pouco estranha (faltam espaços)?}
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \emptyset \vdash M : A}
    {\Gamma ; \emptyset \vdash !M : !A}
\]
\todo{Describe exponent introduction and new gamma context}
\todo{Add h spaces in formula}
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : !A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash let !u = M \mathsf{in} N : C}
\]
\todo{Describe exponent elimination and new gamma context}

@@

Concrete works

\mypara{Type-and-Example-Directed Program Synthesis}~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
In this paper, type synthesis is explored

\item Type-and-example-directed
  synthesis~
  or how to turn your type system upside down.

@@

@@

Mention that you've already implemented a type-checker for this
(useful as a prelim. Exercise but also later, for \emph{validation}).

@@

\[
  \infer[$(\lolli\! I)$]
  {\Delta , x{:}A \vdash M : B}
  {\Delta \vdash \lambda x {:} A . M : A \lolli B}
  \quad
  \infer[$(\lolli\! E)$]
  {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
  {\Delta_1, \Delta_2 \vdash M\,N : B}
\]
